---
title: "Ctr-HPV-coinfection transcriptional modules"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, message=FALSE}
library(limma)
library(data.table)
library(reshape2)
library(pheatmap)

library(ggplot2)

library(clusterProfiler)
library(org.Hs.eg.db)
library(ReactomePA)

result_folder = paste("./Results",format(Sys.time(), "%Y-%m-%d"),sep="/")
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)
```

# Introduction

In this document we will generate summary tables which desribe differences between effects of C.tr alone, HPV E6/E7 alone or coinfection. For each of the conditions, the (co-)infected cells are compared to the non-infected baseline  (ie. condition with no Ctr and no E6E7) or E6E7 transfected cells. 2D and 3D cellular models are analyzed separately. 

For each cellular models, we classifiy each item (gene, gene set or regulon) regarding its result in C.tr alone, coinfection or E6E7 alone in a single class (e.g. Up-Down-Down). Corresponding tables are generated and for genes an enrichment is run for each class. 

```{r}
dge_comparisons = list("2D" = list("Ctr"="B-2D_Ecto_noE6E7_48h_p.i._vs_UI",
                              "HPV"="I_2D_Ecto_UI_E6E7_vs_noE6E7",
                              "CtrHPVCo_vs_E6E7alone"="C-2D_Ecto_E6E7_48h_p.i._vs_UI"), 
                       "3D"=list("Ctr"="E-3D_Ecto_noE6E7_5d_p.i._vs_UI" ,
                                 "HPV"="K_3D_Ecto_UI_E6E7_vs_noE6E7",
                                 "CtrHPVCo_vs_E6E7alone"="F-3D_Ecto_E6E7_5d_p.i._vs_UI"))
```

Those are the contrasts used for 2D and 3D cellular models (names refer to comparisons as defined in differential gene expression documents).

```{r}
tmp = t(as.data.frame(dge_comparisons))
rownames(tmp) = gsub("^X","", rownames(tmp))
colnames(tmp) = "Comparison"
tmp
```


```{r}
dge_env = new.env()
load("./Results/2019-11-19/DGE_analysis_image.Rdata", envir = dge_env)

gsea_env = new.env()
load("./Results/data_store/2019-11-19/fGSEA_Results.Rdata", envir=gsea_env)
load("./Results/data_store/2019-11-19/Geneset_collections.Rdata", envir=gsea_env)

dorothea_env = new.env()
load("./Results/2019-11-19/dorothea_results.Rdata", envir=dorothea_env)
```


# DGE

```{r}
all_results = get("all_results", dge_env)
all_DE_results_tmp = list()
sel_comparisons = unlist(dge_comparisons)

for (nn in names(sel_comparisons)) {
  tc = sel_comparisons[nn]
  tmp = all_results[[tc]]
  tmp$condition = nn
  all_DE_results_tmp[[tc]] = tmp
}
all_DE_results_ts = do.call(rbind, all_DE_results_tmp)

all_DE_results_ts$DE_class = ifelse(all_DE_results_ts$adj.P.Val>0.05, "n.s.", ifelse(all_DE_results_ts$logFC > 0,"Up","Down"))
agg_fun = function(x) ifelse("Up" %in% x & "Down" %in% x, "UpDown", ifelse(!any(x %in% c("Up","Down")), "n.s", paste(unique(x[x!="n.s."]),collapse=";") ) )
#agg_fun = function(x) ifelse("Down" %in% x, "Down",ifelse("Up" %in% x, "Up","n.s."))

all_DE_results_sw = dcast(subset(all_DE_results_ts,grepl("protein_coding",Transcript_Type)), GeneSymbol + EntrezID ~ condition, value.var="DE_class", fun.aggregate=agg_fun)

#all_DE_results_FC_p = dcast.data.table(as.data.table(all_DE_results_ts), GeneSymbol+ProbeName ~ condition, value.var=c("logFC","P.Value","adj.P.Val"))

cols_2D = colnames(all_DE_results_sw)[grepl("2D", colnames(all_DE_results_sw))]
cols_3D = colnames(all_DE_results_sw)[grepl("3D", colnames(all_DE_results_sw))]

all_DE_results_sw$class_2D = apply(all_DE_results_sw[, cols_2D], 1, paste, collapse="_" ) 
all_DE_results_sw$class_3D = apply(all_DE_results_sw[, cols_3D], 1, paste, collapse="_" ) 
all_DE_results_sw$class_equal_2D_3D = (all_DE_results_sw$class_2D == all_DE_results_sw$class_3D)
  
e2gs  = tapply(all_DE_results_sw$GeneSymbol, all_DE_results_sw$EntrezID, paste, collapse=",")

```

## Overview

### 2D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(all_DE_results_sw$class_2D)
```

### 3D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(all_DE_results_sw$class_3D)
```

```{r}
ofile = file.path(result_folder, "DGE_Ctr_HPV_coinfection_2D_3D_classification.txt")
write.table(all_DE_results_sw, ofile,sep="\t", row.names=F, quote=F)
```

Table was written to `r ofile`.

For further analysis, we exclude all subclasses that have less than or equal to 20 genes and all that have conflicting (UpDown) results for probes within the same gene. 

```{r}
genes_by_class_2D = with(subset(all_DE_results_sw, !grepl("UpDown",class_2D) & !is.na(EntrezID)), split(EntrezID, class_2D) )
genes_by_class_2D = genes_by_class_2D[lapply(genes_by_class_2D, length)>20]
genes_by_class_3D = with(subset(all_DE_results_sw, !grepl("UpDown",class_3D) & !is.na(EntrezID)), split(EntrezID, class_3D) )
genes_by_class_3D = genes_by_class_3D[lapply(genes_by_class_3D, length)>20]
```

## ClusterProfiler enrichment

### 2D

#### Enrichment GO 

```{r, fig.height=13, fig.width=14}

ck_go <- compareCluster(geneCluster = genes_by_class_2D, fun = "enrichGO", OrgDb="org.Hs.eg.db", ont="BP")
dotplot(ck_go) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#### Enrichment KEGG

```{r, fig.height=13, fig.width=14}

ck_kegg <- compareCluster(geneCluster = genes_by_class_2D, fun = "enrichKEGG")
dotplot(ck_kegg, showCategory = 15) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#### Enrichment Reactome

```{r, fig.height=12, fig.width=18}
ck_reactome <- compareCluster(geneCluster = genes_by_class_2D, fun = "enrichPathway")
dotplot(ck_reactome) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
analysis_name = "2D"

ck_go@compareClusterResult$database = "GO"
ck_kegg@compareClusterResult$database="KEGG"
ck_reactome@compareClusterResult$database = "REACTOME"

HM_cluster_enrichment = rbind(ck_go@compareClusterResult, ck_kegg@compareClusterResult, ck_reactome@compareClusterResult)


HM_cluster_enrichment$GeneSymbols = gs = Map(function(x) {sapply(strsplit(x,"/"), function(y) sort(paste(e2gs[y], collapse=",")))}, HM_cluster_enrichment$geneID )
cc = colnames(HM_cluster_enrichment)
sel_cols = c("Cluster","database")
sel_cols = append(sel_cols, cc[!cc %in% sel_cols & !cc =="geneID"])

ofile = file.path(result_folder,paste0("DGE_",analysis_name,"_enrichment.txt"))
fwrite(HM_cluster_enrichment[, sel_cols], file=ofile, sep="\t", quote=F)

```

Enrichment results were written to `r ofile`

### 3D

#### Enrichment GO 

```{r, fig.height=12, fig.width=14}

ck_go <- compareCluster(geneCluster = genes_by_class_3D, fun = "enrichGO", OrgDb="org.Hs.eg.db", ont="BP")
dotplot(ck_go) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#### Enrichment KEGG

```{r, fig.height=12, fig.width=14}

ck_kegg <- compareCluster(geneCluster = genes_by_class_3D, fun = "enrichKEGG")
dotplot(ck_kegg, showCategory = 15) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#### Enrichment Reactome

```{r, fig.height=12, fig.width=18}
ck_reactome <- compareCluster(geneCluster = genes_by_class_3D, fun = "enrichPathway")
dotplot(ck_reactome) + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
analysis_name = "3D"

ck_go@compareClusterResult$database = "GO"
ck_kegg@compareClusterResult$database="KEGG"
ck_reactome@compareClusterResult$database = "REACTOME"

HM_cluster_enrichment = rbind(ck_go@compareClusterResult, ck_kegg@compareClusterResult, ck_reactome@compareClusterResult)


HM_cluster_enrichment$GeneSymbols = gs = Map(function(x) {sapply(strsplit(x,"/"), function(y) sort(paste(e2gs[y], collapse=",")))}, HM_cluster_enrichment$geneID )
cc = colnames(HM_cluster_enrichment)
sel_cols = c("Cluster","database")
sel_cols = append(sel_cols, cc[!cc %in% sel_cols & !cc =="geneID"])

ofile = file.path(result_folder,paste0("DGE_",analysis_name,"_enrichment.txt"))
fwrite(HM_cluster_enrichment[, sel_cols], file=ofile, sep="\t", quote=F)
```

Enrichment results were written to `r ofile`


# GSEA

```{r}
all_gsea_results = get("all_gsea_results",gsea_env)

all_gsea_results_tmp = list()
sel_comparisons = unlist(dge_comparisons)

for (nn in names(sel_comparisons)) {
  tc = sel_comparisons[nn]
  tmp = all_gsea_results[[tc]]
  tmp$comparison = nn
  all_gsea_results_tmp[[tc]] = tmp
}
all_gsea_results_ts = do.call(rbind, all_gsea_results_tmp)

all_gsea_results_ts$gsea_class = ifelse(all_gsea_results_ts$FDR_global > 0.05, "n.s.", ifelse(all_gsea_results_ts$NES > 0,"Up","Down"))

#agg_fun = function(x) ifelse("Up" %in% x & "Down" %in% x, "UpDown", ifelse(!any(x %in% c("Up","Down")), "n.s", paste(unique(x[x!="n.s."]),collapse=";") ) )
#agg_fun = function(x) ifelse("Down" %in% x, "Down",ifelse("Up" %in% x, "Up","n.s."))

all_gsea_results_sw = dcast(all_gsea_results_ts, GeneSetCollection + pathway ~ comparison, value.var="gsea_class")

all_gsea_results_sw = all_gsea_results_sw[apply(all_gsea_results_sw[, names(sel_comparisons)]=="n.s.", 1, sum)<length(sel_comparisons),]

cols_2D = colnames(all_gsea_results_sw)[grepl("2D", colnames(all_gsea_results_sw))]
cols_3D = colnames(all_gsea_results_sw)[grepl("3D", colnames(all_gsea_results_sw))]

all_gsea_results_sw$class_2D = apply(all_gsea_results_sw[, cols_2D], 1, paste, collapse="_" ) 
all_gsea_results_sw$class_3D = apply(all_gsea_results_sw[, cols_3D], 1, paste, collapse="_" ) 
all_gsea_results_sw$class_equal_2D_3D = (all_gsea_results_sw$class_2D == all_gsea_results_sw$class_3D)

```

## Overview

### 2D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(all_gsea_results_sw$class_2D)
```

### 3D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(all_gsea_results_sw$class_3D)
```


```{r}
ofile = file.path(result_folder, "GSEA_Ctr_HPV_coinfection_2D_3D_classification.txt")
write.table(all_gsea_results_sw, ofile,sep="\t", row.names=F, quote=F)

```

Table was written to `r ofile`.

# Dorothea

```{r}
all_dorothea_results_complete = get("all_dorothea_results",dorothea_env)
dorothea_res_tabs = list()

for (aa in names(all_dorothea_results_complete)) {
  
  all_dorothea_results = all_dorothea_results_complete[[aa]]
  
  all_dorothea_results_tmp = list()
  sel_comparisons = unlist(dge_comparisons)
  
  for (nn in names(sel_comparisons)) {
    tc = sel_comparisons[nn]
    tmp = all_dorothea_results[[tc]]
    tmp$comparison = nn
    all_dorothea_results_tmp[[tc]] = tmp
  }
  all_dorothea_results_ts = do.call(rbind, all_dorothea_results_tmp)
  
  all_dorothea_results_ts$dorothea_class = ifelse(all_dorothea_results_ts$FDR > 0.05, "n.s.", ifelse(all_dorothea_results_ts$NES > 0,"Up","Down"))
  
  all_dorothea_results_sw = dcast(all_dorothea_results_ts, Regulon ~ comparison, value.var="dorothea_class")
  
  all_dorothea_results_sw = all_dorothea_results_sw[apply(all_dorothea_results_sw[, names(sel_comparisons)]=="n.s.", 1, sum)<length(sel_comparisons),]
  
  cols_2D = colnames(all_dorothea_results_sw)[grepl("2D", colnames(all_dorothea_results_sw))]
  cols_3D = colnames(all_dorothea_results_sw)[grepl("3D", colnames(all_dorothea_results_sw))]
  
  all_dorothea_results_sw$class_2D = apply(all_dorothea_results_sw[, cols_2D], 1, paste, collapse="_" ) 
  all_dorothea_results_sw$class_3D = apply(all_dorothea_results_sw[, cols_3D], 1, paste, collapse="_" ) 
  all_dorothea_results_sw$class_equal_2D_3D = (all_dorothea_results_sw$class_2D == all_dorothea_results_sw$class_3D)
  
  dorothea_res_tabs[[aa]] = all_dorothea_results_sw
  
  ofile = file.path(result_folder, paste0("Dorothea_",aa,"_Ctr_HPV_coinfection_2D_3D_classification.txt"))
  cat(paste(ofile,"\n"))
  write.table(all_dorothea_results_sw, ofile,sep="\t", row.names=F, quote=F)
}

```


## Overview for Level B Dorothea results

### 2D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(dorothea_res_tabs[["LevelB"]]$class_2D)
```

```{r, fig.height=16}
tmp = as.matrix(dorothea_res_tabs[["LevelB"]][,cols_2D])
tmp = ifelse(tmp=="Up",1,ifelse(tmp=="Down",-1,0))
rownames(tmp) = dorothea_res_tabs[["LevelB"]]$Regulon

row_anno = dorothea_res_tabs[["LevelB"]][, "class_2D", drop=F]
rownames(row_anno) = dorothea_res_tabs[["LevelB"]]$Regulon

pheatmap(tmp, annotation_row = row_anno, cluster_cols = F, main="Dorothea results 2D", fontsize_col = 20)
```

### 3D classes

Classes are C.tr infection alone - Coinfection - HPV E6/E7 alone

```{r}
table(dorothea_res_tabs[["LevelB"]]$class_3D)
```

```{r, fig.height=16}
tmp = as.matrix(dorothea_res_tabs[["LevelB"]][,cols_3D])
tmp = ifelse(tmp=="Up",1,ifelse(tmp=="Down",-1,0))
rownames(tmp) = dorothea_res_tabs[["LevelB"]]$Regulon

row_anno = dorothea_res_tabs[["LevelB"]][, "class_3D", drop=F]
rownames(row_anno) = dorothea_res_tabs[["LevelB"]]$Regulon

pheatmap(tmp, annotation_row = row_anno, cluster_cols = F, main="Dorothea results 3D", fontsize_col = 20)
```

# Software versions

```{r}
sessionInfo()
```

